\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-09-23 submission/manuscript.Rmd
intro pass

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Visual illusions highlight how the brain uses contextual and prior information to inform our perception of reality. Unfortunately, illusion research has been hampered by the difficulty of adapting these stimuli to experimental settings. In this study, we used the novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and reaction times, and extracted participant-level performance scores (n=250). Our results provide evidence in favour of a general factor (labelled Factor *i*) underlying the sensitivity to different illusions. Moreover, we report a positive relationship between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "3808"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

<!-- Intro  & Gaps-->

@@

Notably, the presence of a common mechanism underlying the effects of different illusions has been contested [@hamburger2016; @cretenoud2019; @cretenoud2020illusions]. Indeed, while early research had suggested a common factor of illusion sensitivity indexed by overall vision proficiency [@halpern1999interindividual; @thurstone1944factorial], recent studies attempting to identify said factor have at most observed weak correlations between resistance to distinct illusions [@grzeczkowski2017; @grzeczkowski2018]. To this end, the nature of the processes underlying illusion perception - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences of prior beliefs [@teufel2018; @caporuscio2022] remain debated. To add, the existence of dispositional correlates of illusion sensitivity has also been another area of controversy, with some studies reporting higher illusion resistance in individuals with stronger aggression and narcissism traits [@zhang2017; @konrath2009seeing]. In particular, an increasing amount of evidence in recent years purports lower illusion susceptibility in patients with schizophrenia and autism [@notredame2014; @park2022; @grzeczkowski2018; @gori2016; @razeghi2022].

<!-- Bayes -->

In an attempt to bridge the research gaps aforementioned, one recent theory proposes to conceptualize illusions under the Bayesian brain hypothesis [@friston2010free]; as ambiguous percepts (noisy sensory evidence) giving ample weight to prior knowledge to provide a coherent perceptual experience. In this framework, certain traits or characteristics (e.g., psychoticism) are driven by alterations in the system's metacognitive components [@adams2013computational], resulting in underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions [@koethe2009binocular].

@@

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge to empirical efforts involves the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability [@cretenoud2020illusions; @cretenoud2019z; @bressan2021most]. Moreover, conventional paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different [@lanyi2022can], or having them adjust the targets to perceptually match a reference stimulus [@mylniec2016; @grzeczkowski2018]. However, this reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity [@skottun2014subjective].

@@

A pilot study (*n = 46*), of which full description is available in the Supplementary Materials, was first conducted to determine a sensitive range of stimuli parameters. Then, for each of the 10 illusion types, we generated a total of 134 stimuli. These stimuli resulted from the combination of 15 equally-spaced levels of illusion *strength* (7 negative, i.e., congruent effects; 7 positive, i.e., incongruent effects; and 0) overlapped with 16 non-linearly spaced task *difficulty* levels (i.e., with an exponential, square or cubic spacing depending on the pilot results). For instance, a linear space of [0.1, 0.4, 0.7, 1.0] can be transformed to an exponential space of [0.1, 0.34, 0.64, 1.0], where 0.1 corresponds to the highest difficulty - i.e., the smallest objective difference between targets). For each illusion type, the stimuli were split into two series (56 and 72 stimuli per series) with alternating parameter values to maintain their homogeneity. Additionally, 6 stimuli per illusion type was generated for a practice series using parameters with more extreme variations (i.e., containing very easy trials to help cement the task instructions).

>>>>

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Visual illusions highlight how the brain uses contextual and prior information to inform our perception of reality. Unfortunately, illusion research has been hampered by the difficulty of adapting these stimuli to experimental settings. In this study, we used the novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and reaction times, and extracted participant-level performance scores (n=250). Our results provide evidence in favour of a general factor (labelled Factor *i*) underlying the sensitivity to different illusions. Moreover, we report a positive relationship between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "3876"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

<!-- Intro -->

@@

<!-- Gaps-->

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions [@hamburger2016; @cretenoud2020illusions]. While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency [@halpern1999interindividual; @thurstone1944factorial], recent empirical studies at most observed weak correlations between resistance to distinct illusions [@grzeczkowski2017; @grzeczkowski2018]. The existence of dispositional correlates of illusion sensitivity has also been controversial, related to evidence suggesting a lower illusion sensitivity in in patients with schizophrenia and autism [@notredame2014; @park2022; @grzeczkowski2018; @gori2016; @razeghi2022], as well as individuals with stronger aggression and narcissism traits [@zhang2017; @konrath2009seeing].

<!-- Bayes -->
Although the nature of the processes underlying illusion perception - whether related to low-level features of the visual processing system [@cretenoud2019; @gori2016] or to top-down influences [@teufel2018; @caporuscio2022] - remains debated, a growing body of literature proposes to conceptualize illusions under the Bayesian brain hypothesis [@friston2010free], as ambiguous percepts (noisy sensory evidence) giving ample weight to prior knowledge to minimize prediction error and provide a coherent perceptual experience. In this framework, certain dispositional traits or characteristics (e.g., psychoticism) are seen as driven by alterations in the system's metacognitive components [@adams2013computational], resulting in an underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions [@koethe2009binocular].

@@

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge can be found in the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability [@cretenoud2020illusions; @cretenoud2019; @bressan2021most]. Moreover, conventional paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different [@lanyi2022can], or having them adjust the targets to perceptually match a reference stimulus [@mylniec2016; @grzeczkowski2018]. This reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity [@skottun2014subjective].

@@

A pilot study (*n = 46*), of which full description is available in the Supplementary Materials, was first conducted to determine a sensitive range of stimuli parameters. Then, for each of the 10 illusion types, we generated a total of 134 stimuli. These stimuli resulted from the combination of 15 equally-spaced levels of illusion *strength* (7 negative, i.e., congruent effects; 7 positive, i.e., incongruent effects; and 0) overlapped with 16 non-linearly spaced task *difficulty* levels (i.e., with an exponential, square or cubic spacing depending on the pilot results). For instance, a linear space of [0.1, 0.4, 0.7, 1.0] can be transformed to an exponential space of [0.1, 0.34, 0.64, 1.0], where 0.1 corresponds to the highest difficulty - i.e., the smallest objective difference between targets). For each illusion type, the stimuli were split into two series (56 and 72 stimuli per series) with alternating parameter values to maintain their homogeneity. Additionally, 6 stimuli per illusion type were generated for a practice series using parameters with more extreme variations (i.e., containing very easy trials to help cement the task instructions).

### 2022-09-23 submission/manuscript.tex
intro pass

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{visual illusions, illusion game, Pyllusion, personality, general factor\newline\indent Word count: 3628}
\usepackage{lineno}

@@

Notably, the presence of a common mechanism underlying the effects of different illusions has been contested (Cretenoud et al., 2019; Cretenoud et al., 2020; Hamburger, 2016); and the nature of the underlying processes - whether related to low-level features of the visual processing system (Cretenoud et al., 2019; Gori et al., 2016) or to top-down influences of prior beliefs (Caporuscio et al., 2022; Teufel et al., 2018) remain debated. The existence of dispositional correlates of illusion sensitivity is another area of controversy, with some studies reporting higher illusion resistance in patients with schizophrenia and autism (Giaouri \& Alevriadou, 2011; Keane et al., 2014; Notredame et al., 2014; Park et al., 2022; Pessoa et al., 2008), and in individuals with stronger aggression and narcissism traits (Konrath et al., 2009; Zhang et al., 2017).

One recent theory proposes to conceptualize illusions under the Bayesian brain hypothesis (Friston, 2010), as ambiguous percepts (noisy sensory evidence) giving ample weight to prior knowledge to provide a coherent perceptual experience. In this framework, certain traits or characteristics (e.g., psychoticism) are driven by alterations in the system's metacognitive components (Adams et al., 2013), resulting in underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions (Koethe et al., 2009).

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge is the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability. Moreover, the paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different (Lányi et al., 2022), or having them adjust the targets to perceptually match a reference stimulus (Grzeczkowski et al., 2018; Mylniec \& Bednarek, 2016). However, this reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity.

@@

A pilot study (\emph{n = 46}), of which full description is available in the Supplementary Materials, was first conducted to determine a sensitive range of stimuli parameters. Then, for each of the 10 illusion types, we generated a total of 134 stimuli. These stimuli resulted from the combination of 15 equally-spaced levels of illusion \emph{strength} (7 negative, i.e., congruent effects; 7 positive, i.e., incongruent effects; and 0) overlapped with 16 non-linearly spaced task \emph{difficulty} levels (i.e., with an exponential, square or cubic spacing depending on the pilot results). For instance, a linear space of {[}0.1, 0.4, 0.7, 1.0{]} can be transformed to an exponential space of {[}0.1, 0.34, 0.64, 1.0{]}, where 0.1 corresponds to the highest difficulty - i.e., the smallest objective difference between targets). For each illusion type, the stimuli were split into two series (56 and 72 stimuli per series) with alternating parameter values to maintain their homogeneity. Additionally, 6 stimuli per illusion type was generated for a practice series, with more extreme variations (i.e., containing very easy trials to help cement the task instructions).

@@

After a brief demographic survey and a practice series of illusions, the first series of 10 illusion blocks was presented in a randomized order, with a further randomization of the stimuli order within each block. Following this first series of blocks, two personality questionnaires were administered, the \emph{IPIP6} (24 items, Sibley et al., 2011) - measuring 6 ``normal'' personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility), and the \emph{PID-5} (25 items, Hopwood et al., 2012) - measuring 5 ``pathological'' personality traits (Disinhibition, Antagonism, Detachment, Negative Affect and Psychoticism). Next, the second series of 10 illusion blocks was presented (with new randomized orders of blocks and trials). In total, each participant underwent 1340 trials of which they had to respond ``as fast as possible without making errors'' (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off) by pressing the correct arrow key (left/right,or up/down depending on the illusion type). For instance, in the Müller-Lyer block, participants had to answer which one of the upper or bottom target line was the longest. The task was implemented using \emph{jsPsych} (De Leeuw, 2015), and the set of instructions for each illusion type is available in the experiment code.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-giaouri2011}{}}%
Giaouri, S., \& Alevriadou, A. (2011). Are children with down syndrome susceptible to visual illusions? \emph{Procedia - Social and Behavioral Sciences}, \emph{15}, 1988--1992. \url{https://doi.org/10.1016/j.sbspro.2011.04.040}

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-keane2014}{}}%
Keane, B. P., Joseph, J., \& Silverstein, S. M. (2014). Late, not early, stages of Kanizsa shape perception are compromised in schizophrenia. \emph{Neuropsychologia}, \emph{56}, 302--311. \url{https://doi.org/10.1016/j.neuropsychologia.2014.02.001}

@@

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{visual illusions, illusion game, Pyllusion, personality, general factor\newline\indent Word count: 3873}
\usepackage{lineno}

@@

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions (Cretenoud et al., 2020; Hamburger, 2016). While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency (Halpern et al., 1999; Thurstone, 1944), recent empirical studies at most observed weak correlations between resistance to distinct illusions (Grzeczkowski et al., 2017, 2018). The existence of dispositional correlates of illusion sensitivity has also been controversial, related to evidence suggesting a lower illusion sensitivity in in patients with schizophrenia and autism (Gori et al., 2016; Grzeczkowski et al., 2018; Notredame et al., 2014; Park et al., 2022; Razeghi et al., 2022), as well as individuals with stronger aggression and narcissism traits (Konrath et al., 2009; Zhang et al., 2017).

Although the nature of the processes underlying illusion perception - whether related to low-level features of the visual processing system (Cretenoud et al., 2019; Gori et al., 2016) or to top-down influences (Caporuscio et al., 2022; Teufel et al., 2018) - remains debated, a growing body of literature proposes to conceptualize illusions under the Bayesian brain hypothesis (Friston, 2010), as ambiguous percepts (noisy sensory evidence) giving ample weight to prior knowledge to minimize prediction error and provide a coherent perceptual experience. In this framework, certain dispositional traits or characteristics (e.g., psychoticism) are seen as driven by alterations in the system's metacognitive components (Adams et al., 2013), resulting in an underweighting of priors during perceptual inferences, and manifesting as a decreased sensitivity to illusions (Koethe et al., 2009).

Despite strong theoretical foundations and hypotheses, the empirical evidence remains scarce, clouded by methodological hurdles. For instance, one key challenge can be found in the difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. Instead, existing studies typically use only one or a small subset of illusion types, with few contrasting conditions, restricting the findings' generalizability (Bressan \& Kramer, 2021; Cretenoud et al., 2019; Cretenoud et al., 2020). Moreover, conventional paradigms often focus on the participants' subjective experience, by asking them the extent to which they perceive two identical targets as different (Lányi et al., 2022), or having them adjust the targets to perceptually match a reference stimulus (Grzeczkowski et al., 2018; Mylniec \& Bednarek, 2016). This reliance on meta-cognitive judgements about one's subjective experience likely distorts the measurand, limiting the ability to reliably obtain more direct and objective measures of illusion sensitivity (Skottun \& Skoyles, 2014).

@@

A pilot study (\emph{n = 46}), of which full description is available in the Supplementary Materials, was first conducted to determine a sensitive range of stimuli parameters. Then, for each of the 10 illusion types, we generated a total of 134 stimuli. These stimuli resulted from the combination of 15 equally-spaced levels of illusion \emph{strength} (7 negative, i.e., congruent effects; 7 positive, i.e., incongruent effects; and 0) overlapped with 16 non-linearly spaced task \emph{difficulty} levels (i.e., with an exponential, square or cubic spacing depending on the pilot results). For instance, a linear space of {[}0.1, 0.4, 0.7, 1.0{]} can be transformed to an exponential space of {[}0.1, 0.34, 0.64, 1.0{]}, where 0.1 corresponds to the highest difficulty - i.e., the smallest objective difference between targets). For each illusion type, the stimuli were split into two series (56 and 72 stimuli per series) with alternating parameter values to maintain their homogeneity. Additionally, 6 stimuli per illusion type were generated for a practice series using parameters with more extreme variations (i.e., containing very easy trials to help cement the task instructions).

@@

After a brief demographic survey and a practice series of illusions, the first series of 10 illusion blocks was presented in a randomized order, with a further randomization of the stimuli order within each block. Following this first series of blocks, two personality questionnaires were administered, the \emph{IPIP6} (24 items, Sibley et al., 2011) - measuring 6 ``normal'' personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility), and the \emph{PID-5} (25 items, Hopwood et al., 2012) - measuring 5 ``pathological'' personality traits (Disinhibition, Antagonism, Detachment, Negative Affect and Psychoticism). Next, the second series of 10 illusion blocks was presented (with new randomized orders of blocks and trials). In total, each participant underwent 1340 trials of which they had to respond ``as fast as possible without making errors'' (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off) by pressing the correct arrow key (left/right, or up/down depending on the illusion type). For instance, in the Müller-Lyer block, participants had to answer which one of the upper or bottom target line was the longest. The task was implemented using \emph{jsPsych} (De Leeuw, 2015), and the set of instructions for each illusion type is available in the experiment code.

@@

\leavevmode\vadjust pre{\hypertarget{ref-bressan2021most}{}}%
Bressan, P., \& Kramer, P. (2021). Most findings obtained with untimed visual illusions are confounded. \emph{Psychological Science}, \emph{32}(8), 1238--1246.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-halpern1999interindividual}{}}%
Halpern, S. D., Andrews, T. J., \& Purves, D. (1999). Interindividual variation in human visual performance. \emph{Journal of Cognitive Neuroscience}, \emph{11}(5), 521--534.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-skottun2014subjective}{}}%
Skottun, B. C., \& Skoyles, J. R. (2014). Subjective criteria and illusions in visual testing: Some methodological limitations. \emph{Psychological Research}, \emph{78}(1), 136--140.

@@

\leavevmode\vadjust pre{\hypertarget{ref-thurstone1944factorial}{}}%
Thurstone, L. L. (1944). \emph{A factorial study of perception.}
